\textbf{Microbiota predict \emph{Clostridium difficile} severity in germ-free mice colonized with human feces}

### 2021-12-14 submission/manuscript.Rmd
update manuscript with pds edits round 2

# The gut bacterial community potentiates *Clostridiodes difficile* infection severity.

@@

*Clostridioides difficile* infection (CDI) disease severity has increased over the last few decades. Patient age, white blood cell count, creatinine levels as well as *C. difficile* ribotype and toxin genes have been associated with disease severity. However, current models of CDI disease severity that incorporate these factors are not robust enough to be broadly applied. These models lack data describing the gut microbiota. Perturbations to the gut microbiota are necessary for *C. difficile* to colonize the gut. The gut microbiota is thought to impair *C. difficile* colonization through bile acid metabolism, nutrient consumption and bacteriocin production. It has been unclear whether the gut microbiota affect CDI disease severity. However, here we demonstrate that gut bacterial communities can contribute to disease severity. We derived diverse gut communities by colonizing germ-free mice with different human fecal communities. The mice were then infected with a single *C. difficile* RT027 clinical isolate which resulted in moribundity and histopathologic differences. The variation in severity was associated with the human fecal community that the mice received. Generally, facultative anaerobes with pathogenic potential, such as *Escherichia*, *Helicobacter*,  and *Klebsiella*, were associated with more severe outcomes. Bacterial groups associated with fiber degradation, such as *Coprobacillus*, were associated with less severe outcomes. Finally, we showed that, when infected with a different *C. difficile* isolate, the outcome of the infection was similar between the mice that received the same fecal donor community. These data have demonstrated that the gut bacteria can influence CDI disease severity and may improve clinical models.

@@

*Clostridioides difficile* infections (CDI) can be asymptomatic or range in severity from mild diarrhea to severe outcomes including surgical interventions and death. Models that predict severity and guide treatment decisions are based on clinical factors and *C. difficile* characteristics. Although the gut microbiome plays a role in protecting against CDI, it has not been investigated for its effect on CDI disease severity or incorporated into attempts to predict disease severity. We demonstrated that variation in the microbiome of mice colonized with human feces was sufficient to result in a range of disease outcomes. These results revealed groups of bacteria associated with both severe and mild *C. difficile* infection outcomes. Gut bacterial community data from patients with CDI could improve our ability to identify patients at risk of developing more severe disease and improve interventions which treat the gut bacteria to reduce host damage.

@@

*Clostridioides difficile* infections (CDI) have increased in incidence and severity since *C. difficile* was first identified as the cause of antibiotic-associated pseudomembranous colitis [@Kelly2008]. CDI disease severity can range from asymptomatic to mild diarrhea to toxic megacolon and death. Treatment for CDI is driven by the disease severity. The Infectious Diseases Society of America (IDSA) and Society for Healthcare Epidemiology of America (SHEA) guidelines categorize severe disease based on white blood cell count greater than 15,000 cells/mm$^3$ and/or a serum creatinine greater than 1.5 mg/dL, and complicated disease based on systolic blood pressure below 90, ileus, or toxic megacolon [@McDonald2018]. Since these measures are only present once the disease has advanced, the IDSA classification does not allow for identification of patients at risk of severe CDI. Thus, we have limited ability to prevent patients from developing severe CDI. Ideally, we could identify patients that are high-risk of severe disease and adjust their treatments to halt the disease before it worsens.

Prediction models have been developed to score a patients risk for severe CDI outcomes but have not been robust for broad application [@Dieterle2020; @Butt2013; @Perry2021; @vanBeurden2017]. There is no widely accepted definition of severe disease [@Ressler2021]. While the IDSA guideline is frequently used, many studies develop their own severity definitions. Aside from the features utilized by the IDSA guidelines, other severity measures use patient demographics, vital information, laboratory test results, disease complications, or infection outcomes such as ICU admission, surgery and/or death [@Dieterle2020; @Zar2007; @Gujja2009; @Jardin2013]. However, a recent multi-center external validation of these scoring systems using a consistent severity definition - transfer or admission to the intensive care unit, colectomy, and/or death within 30 days of diagnosis - demonstrated a decline in performance and high false positivity rate [@Perry2021]. Thus, patient and clinical factors have not been sufficient to accurately identify patients at risk of severe CDI and may benefit from additional information involved in the development of CDI disease.

Severity has been associated with characteristics of *C. difficile* ribotypes, such as 027, and host factors [@Bauer2011]. Ribotypes with genes for toxin TcdA/TcdB have the potential to disrupt the tight junctions between the intestinal epithelial cells [@Abt2016]. The presence of binary toxin CDT or mutations in *tcdC*, the negative regulator of toxin transcription, have been associated with increased virulence of TcdA/TcdB [@Abt2016]. In addition to the pathogenic potential of *C. difficile*, the response by the host also affects the disease severity. IgG antibody activity towards the TcdA/TcdB are decreased in both patients with symptomatic and recurrent infections [@Kelly2011]. In addition to controlling the infection, the immune response can also cause intestinal injury. Increased neutrophil inflitration into the mucosa after the cellular tight junctions have been compromised can exacerbate the inflammatory response [@Farooq2015]. While we have have elucidated these major drivers of host damage in CDI, these are also the drivers of features of the poorly performing prediction models.

Missing from CDI severity prediction models are the effects of the indigenous gut bacteria. *C. difficile* interacts with the gut community in many ways. The indigenous bacteria of a healthy intestinal community provide a protective barrier preventing *C. difficile* from infecting the gut. A range of mechanisms can disrupt this barrier, including antibiotics, medications, or dietary changes, and lead to increased susceptiblity to CDI [@Britton2014; @Hryckowian2018; @VichVila2020]. Once in the gut, the indigenous bacteria can either promote or inhibit *C. difficile* through producing molecules or modifying the environment [@Abbas2020]. Bile acids metabolized by the gut bacteria can inhibit *C. difficile* growth and affect toxin production [@Sorg2008; @Thanissery2017]. Bacteria in the gut also can compete more directly with *C. difficile* through antibiotic production or nutrient consumption [@Aguirre2021; @Kang2019]. While the relationship between the gut bacteria and *C. difficile* has been established, the effect the gut bacteria can have on CDI disease severity is unclear. Recent studies have begun to uncover this relationship but have either used clinical data, which does not permit control of the host or *C. difficile* variation [@Seekatz2016] or murine experiments, which lack variation in the bacterial communities [@Dieterle2020]. Recently, Tomkovich et al. showed the variation in the bacterial communities between mice from different mouse colonies resulted in different clearance rates of *C. difficile* [@Tomkovich2020]. Therefore, it is clear the gut bacteria modulate the activity and infection dynmaics of *C. difficile*. We hypothesized that gut bacteria also contribute to the variation in CDI disease severity. Here, we colonized germ-free C57BL/6 mice with human fecal communities to test whether the variability in the gut bacteria can produce and explain variation in disease severity.

@@

***C. difficile* is able to infect germ-free mice colonized with human fecal microbial communities without perturbation.**  To produce gut microbiomes with greater variation than those found in conventional mouse colonies, we colonized germ-free mice with bacteria from human feces [@Nagpal2018]. We inoculated germ-free C57BL/6 mice with homogenized feces from one of 15 human fecal samples via oral gavage. The human fecal samples were selected because they represented diverse community structures based on community clustering [@Schubert2014]. The gut communities were allowed to establish themselves for two weeks post inoculation [@Gillilland2012]. We then surveyed the bacterial members of the gut communities by 16S rRNA gene sequencing of fecal pellets (Figure 1A). The bacterial communities from each mouse grouped more closely to those communities from mice that received the same human fecal donor community than to the mice who received a different human fecal donor community (Figure 1B). The communities were primarily composed of populations of *Clostridia*, *Bacteroidia*, *Erysipelotrichia*, *Bacilli*, and *Gammaproteobacteria*. However, each group of mice harbored unique combination of class relative abundances in their gut bacterial communities.

Next, we tested this set of mice with their human-derived gut microbial communities for susceptibility to *C. difficile* infection. A typical mouse model of CDI requires pre-treatment with antibiotics, such as clindamycin, to become susceptible to *C. difficile* colonization [@Chen2008; @Schubert2015]. However, we wanted to avoid modifying the communities with an antibiotic to maintain their unique microbial compositions and ecological relationships. Since these communities came from people who have increased risk of CDI, such as hospitalization or recent antibiotic use [@Schubert2014], we decided to test if *C. difficile* was able to infect these mice without any antibiotic perturbation. We hypothesized that *C. difficile* would be able to colonize a subset of mice whose communities were susceptible from perturbations either the donor experienced or from the ecological selection when colonizing the murine gut. Mice were challenged with 10$^{3}$ *C. difficile* ribotype 027 human isolate 431 spores. The mice were followed for 10 days post-challenge and their stool was collected and plated for *C. difficile* colony forming units (CFU) to determine the extent of the infection. Suprisingly, mice from all donors were colonized (Figure 2). Only two mice were able to resist *C. difficile* infection and both received their community from the same donor, N3. By colonizing germ-free mice with different human fecal communities, we were able to generate diverse gut communities in mice, which were susceptible to *C. difficile* infection without further modification of the gut community.

**Infection severity varies by initial community.** After we challenged the mice with *C. difficile*, we investigated the outcome from the infection and its relationship to the initial community. We followed the mice for 10 days post-challenge for colonization density, toxin production, and mortality. Seven mice, from Donors N1, N2 and N5, were not colonized on the day after *C. difficile* challenge but were infected (>10$^{6}$) by the end of the experiment. All mice that received their community from Donor M1 through M6 succumbed to the infection and became moribund within 3 days post-challenge. The remaining mice, except the uninfected Donor N3 mice, maintained *C. difficile* infection through the end of the experiment (Figure 2). At 10 days post-challenge, or earlier for the moribund mice, mice were euthanised and fecal pellets were assayed for toxin activity and cecal tissue was collected and scored for histopathologic signs of disease (Figure 3). Overall, there was greater toxin activity detected in the stool of the moribund mice (*P* = 0.003). However, when looking at each group of mice, we saw there was a range in toxin activity for both the moribund and non-moribund mice (Figure 3A). Non-moribund mice from Donors N4 through N9 had comparable toxin activity as the moribund mice. Additionally, not all moribund mice had toxin activity detected in their stool. This discordance between toxin activity and disease severity has been previously described [@Akerlund2006; @Vitucci2020]. Next we examined the cecal tissue for histopathologic damage. Moribund mice had high levels of epithelial damage, tissue edema and inflammation (Figure S1), similar to previously reported histopathologic findings for *C. difficile* ribotype 027 [@Cowardin2016]. As observed with toxin activity, the moribund mice had higher histopathologic scores than the non-moribund mice (*P* = 3.0e-9). However, unlike the toxin activity, all moribund mice had consistently high histopathologic summary scores (Figure 3B). The non-moribund mice, Donor groups N1 through N9, had a range in tissue damage from none detected to similar levels as the moribund mice, which grouped by community donor. Together, the toxin activity, histopathologic score, and moribundity showed variation across the donor groups but were largely consistent within each donor group.

**Microbial community members explain variation in CDI severity.** We next interrogated the bacterial communities at the time of *C. difficile* challenge (day 0) for their relationship to infection outcomes using linear discriminant analysis (LDA) effect size (LEfSe) analysis to identify individual bacterial populations that could explain the variation in disease severity. We split the mice into groups by severity level based on their moribundity and histopathologic score. We dichotomized the histopathologic scores into high and low groups by splitting on the median score. This analysis revealed 20 genera that were significantly different by the disease severity (Figure 4A). Bacterial genera *Turicibacter*, *Streptococcus*, *Staphylococcus*, *Pseudomonas*, *Phocaeicola*, *Parabacteroides*, *Bacteroides*, and *Escherichia/Shigella* were detected at higher relative abundances in the mice that became moribund. Populations of *Anaerotignum*, *Coprobacillus*, *Enterocloster*, and *Murimonas* were more abundant in the non-moribund mice that would develop only low intestinal injury. To understand the role of toxin activity in disease severity, we applied LEfSe to identify the genera most likely to explain the observed differences in toxin activity (Figure 4B). Many genera that associated with toxin production were also associated with moribundity, such as populations of *Escherichia/Shigella* and *Bacteroides*. Likewise, there were genera such as *Anaerotignum*, *Enterocloster*, and *Murimonas* that associated with no toxin that also exhibited greater relative abundance in communities from non-moribund mice with a low histopathologic score. Lastly, we tested for correlations between the endpoint relative abundances of bacterial operational taxonomic units (OTUs) and the histopathologic summary score (Figure 4C). The endpoint relative abundance of *Bacteroides* was positively correlated with histopathologic score, as its day 0 relative abundance did with disease severity (Figure 4A). Populations of *Klebsiella* and *Prevotellaceae* were positively correlated with the histopathologic score and were increased in the group of mice with detectable toxin. This analysis has revealed bacterial genera that associated with the variation in moribundity, histopathologic score, and toxin activity of CDI.

We next determined whether collectively, bacterial community membership and relative abundance could be predictive of the CDI disease outcome. We trained random forest models with bacterial community relative abundance data from the day of colonization at each taxonomic rank to predict toxin production, moribundity, and day 10 post-challenge histopathologic summary score. For predicting if toxin would be produced, microbial populations aggregated by phylum rank classification performed similarly as models using lower taxonomic ranks (AUROC = 0.83, Figure S2). *C. difficile* was more likely to produce toxin when the community it infected had less abundant populations of *Verrucomicrobia* and *Campilobacterota* and had more abundant populations of *Proteobacteria* (Figure 5A). Next, we assessed the ability of the community to predict moribundity. Bacteria grouped by class rank classification was sufficient to predict which mice would succumb to the infection before the end of the experiment (AUROC = 0.91, Figure S2). THe features with the greatest effect showed that communities with greater populations of bacteria belonging to *Bacilli* and *Firmicutes* and reduced populations of *Erysipelotrichia* were more likely to result in moribundity (Figure 5B). Only one other class of bacteria was decreased in moribund mice, a group of unclassified *Clostridia*. Lastly, the relative abundances of genera were able to predict a high or low histopathologic score (histopathologic scores were dichotomized as in previous analysis, AUROC = 0.99, Figure S2). No genera had a significantly greater effect on the model performance than any others, indicating the model is reliant on many genera for the correct prediction. The model used some of the genera identified in the LEfSe analysis, such as *Coprobacillus*, *Anaerostipes*, and *Hungatella*. Communities with greater abundances of *Hungatella*, *Eggerthella*, *Bifidobacterium*, *Duncaniella* and *Neisseria* were more likely to have high histopathologic scores. These models have shown that the relative abundance of bacterial populations and their relationship to each other could be used to predict the variation in moribundity, histopathologic score, and toxin activity of CDI.

**Disease severity consistent by donor community across isolates.** Since Carlson et al. showed clinical disease severity varied by *C. difficile* isolate characteristics, such as germination rate [@Carlson2013; @Carlson2015], we next tested whether the effect of the community on CDI outcomes was conserved across *C. difficile* isolates. We colonized mice with human fecal slurry of either Donor N7, N6, or M4 and allowed two weeks for the community to establish. We then challenged the mice with a different *C. difficile* isolate and all mice were infected. The moribundity and histopathologic score outcomes when challenged with a different *C. difficile* human isolate were comparable to the outcomes of challenges with *C. difficile* ribotype 027 human isolate 431 by their donor group (Figure 6). For the isolates and communities we tested, the effect of the gut bacterial community on disease severity was greater than the effect of the variation in these isolates.

@@

Challenging mice colonized with different human fecal communities with a single *C. difficile* isolate, ribotype 027 isolate 431, demonstrated the variation in the gut bacteria affects the *C. difficile* infection disease severity. Our analysis revealed an association between the relative abundance of bacterial community members and disease severity. We used the relative abundance patterns of community members associated with disease outcomes to select donor communities to test whether the disease severity was conserved when challenged with a different *C. difficile* isolate. The disease outcomes from the challenge with the second *C. difficile* isolate matched the outcomes of the challenge with *C. difficile* ribotype 027 isolate 431. Therefore, the effect of the community variation on disease severity *in vivo* was greater than the effect of *C. difficile* isolate variation [@Carlson2013; @Carlson2015]. Overall, these results have demonstrated that the bacterial community contributed to the disease severity of *C. difficile* infection.

CDI severity can range from no symptoms to mild symptoms, such as diarrhea, to severe symptoms, such as toxic megacolon or death. Physicians use prediction tools to identify patients most at risk of developing a severe infection using white blood cell counts, serum albumin level, or serum creatinine level [@Lungulescu2011; @McDonald2018; @Zar2007]. Those levels are driven by the activities in the intestine [@diMasi2018]. Research into the drivers of this variation have revealed factors that make *C. difficile* more virulent. Strains are categorized for their virulence by the presence and production toxin TcdA, toxin TcdB, and binary toxin and the prevelance in outbreaks, such as ribotypes 027 and 078 [@AbernathyClose2020; @Theriot2011; @Goorhuis2008; @OConnor2009; @Chen2008]. However, other studies have shown that virulence is not necessarily linked with toxin production [@Rao2015] or the strain [@Walk2012]. Furthermore, there is variation in the genome, growth rate, sporulation, germination, and toxin production in different isolates of a strain [@Carlson2013; @Carlson2015; @He2010]. This variation may help explain why severe CDI prediction tools often miss identifying many patients with CDI that will develop severe disease [@Dieterle2020; @Butt2013; @Perry2021; @vanBeurden2017]. Therefore, it is necessary to gain a full understanding of all factors contributing to disease variation to improve our ability to predict severity.

The state of the gut bacterial community determines the ability of *C. difficile* to colonize, persist in the intestine. *C. difficile* is unable to colonize an unperturbed healthy gut community and is only able to become established after a perturbation [@Schubert2015]. Once colonized, the different communities lead to different metabolic responses and dynamics of the *C. difficile* population [@Jenior2017; @Jenior2018; @Lesniak2021]. The gut bacteria metabolize primary bile acids into secondary bile acids [@Staley2016; @Long2017]. The concentration of these bile acids affect the germination, growth, toxin production and biofilm formation [@Sorg2008; @Sorg2010; @Thanissery2017; @Dubois2019]. Members of the bacterial community also affect other metabolites *C. difficile* utilizes. *Bacteroides thetaiotaomicron* produce sialidases which release sialic acid from the mucosa for *C. difficile* to utilize [@Ng2013; @Ferreyra2014]. The nutrient environment affects toxin production [@MartinVerstraete2016]. Thus, many of the actions of the gut bacteria modulate *C. difficile* in ways that could affect the infection and resultant disease.

>>>>

# The gut bacterial community potentiates *Clostridioides difficile* infection severity.

@@

The severity of *Clostridioides difficile* infections (CDI) has increased over the last few decades. Patient age, white blood cell count, creatinine levels as well as *C. difficile* ribotype and toxin genes have been associated with disease severity. However, it is unclear whether there is an associate between the gut microbiota disease severity. The gut microbiota is known to interact with *C. difficile* during infection. Perturbations to the gut microbiota are necessary for *C. difficile* to colonize the gut. The gut microbiota is thought to impair *C. difficile* colonization through bile acid metabolism, nutrient consumption and bacteriocin production. It has not been shown whether the gut microbiota can affect CDI disease severity. Here we demonstrate that gut bacterial communities can contribute to disease severity. We derived diverse gut communities by colonizing germ-free mice with different human fecal communities. The mice were then infected with a single *C. difficile* ribotype 027 clinical isolate which resulted in moribundity and histopathologic differences. The variation in severity was associated with the human fecal community that the mice received. Generally, facultative anaerobes with pathogenic potential, such as *Escherichia*, *Helicobacter*,  and *Klebsiella*, were associated with more severe outcomes. Bacterial groups associated with fiber degradation, such as *Coprobacillus*, were associated with less severe outcomes. Finally, we showed that, when infected with a different *C. difficile* isolate, the outcome of the infection was similar between the mice that received the same fecal donor community. These data indicate that gut bacteria can influence CDI disease severity.

@@

*Clostridioides difficile* infections (CDI) can be asymptomatic or range in severity from mild diarrhea to severe outcomes including surgical interventions and death. Models that predict severity and guide treatment decisions are based on clinical factors and *C. difficile* characteristics. Although the gut microbiome plays a role in protecting against CDI, it has not been investigated for its effect on CDI disease severity or incorporated into attempts to predict disease severity. We demonstrated that variation in the microbiome of mice colonized with human feces yielded a range of disease outcomes. These results revealed groups of bacteria associated with both severe and mild *C. difficile* infection outcomes. Gut bacterial community data from patients with CDI could improve our ability to identify patients at risk of developing more severe disease and improve interventions which treat the gut bacteria to reduce host damage.

@@

*Clostridioides difficile* infections (CDI) have increased in incidence and severity since *C. difficile* was first identified as the cause of antibiotic-associated pseudomembranous colitis [@Kelly2008]. CDI disease severity can range from asymptomatic to mild diarrhea to toxic megacolon and death. The Infectious Diseases Society of America (IDSA) and Society for Healthcare Epidemiology of America (SHEA) guidelines categorize severe disease based on white blood cell count greater than 15,000 cells/mm$^3$ and/or a serum creatinine greater than 1.5 mg/dL, and complicated disease based on systolic blood pressure below 90, ileus, or toxic megacolon [@McDonald2018]. Since these measures are only present once the disease has advanced, the IDSA classification does not allow for identification of patients at risk of severe CDI. Prediction models have been developed to score a patient's risk for severe CDI outcomes based on clinical factors but have not been robust for broad application [@Dieterle2020; @Butt2013; @Perry2021; @vanBeurden2017]. Thus, we have limited ability to prevent patients from developing severe CDI. Ideally, we could identify patients that are high-risk of severe disease so their treatments could be adjusted to halt the infection before the disease worsens.

Missing from CDI severity prediction models are the effects of the indigenous gut bacteria. *C. difficile* interacts with the gut community in many ways. The indigenous bacteria of a healthy intestinal community provide a protective barrier preventing *C. difficile* from infecting the gut. A range of mechanisms can disrupt this barrier, including antibiotics, medications, or dietary changes, and lead to increased susceptibility to CDI [@Britton2014; @Hryckowian2018; @VichVila2020]. Once in the gut, the indigenous bacteria can either promote or inhibit *C. difficile* through producing molecules or modifying the environment [@Abbas2020; @Jenior2017]. Bile acids metabolized by the gut bacteria can inhibit *C. difficile* growth and affect toxin production [@Sorg2008; @Thanissery2017]. Bacteria in the gut also can compete more directly with *C. difficile* through antibiotic production or nutrient consumption [@Aguirre2021; @Kang2019]. While the relationship between the gut bacteria and *C. difficile* has been established, the effect the gut bacteria can have on CDI disease severity is unclear.

Studies investigating the severity of CDI disease involving the microbiome have not shown clear evidence of a relationship between the microbiome and disease severity. These studies were not setup to identify a relationship between the microbiome and disease severity. Studies that have used clinical data cannot control for host, microbiome or *C. difficile* ribotype variation [@Seekatz2016]. While disease severity varies, it cannot be attributed to specifically the microbiome since all factors vary. Murine experiments typically use a single mouse colony and different *C. difficile* ribotypes to create severity differences [@Dieterle2020]. Recently, our group has begun uncovering the effect microbiome variation has on *C. difficile* infection. We showed the variation in the bacterial communities between mice from different mouse colonies resulted in different clearance rates of *C. difficile* [@Tomkovich2020]. We also showed varied ability of mice to spontaneously eliminate *C. difficile* infection when they were treated with different antibiotics prior to *C. difficile* challenge [@Lesniak2021]. However, there have not been any studies investigating whether variation in the microbiome affects the resultant disease severity of *C. difficile* infections.

Here, we sought to determine whether the gut bacteria contribute to disease severity. We hypothesized that since the gut bacteria affect the metabolism of *C. difficile* and its infection dynamics, it can also affect the disease severity of the infection. To test this hypothesis, we colonized germ-free C57BL/6 mice with human fecal communities to create varied gut communities. We then challenged the mice with a single *C. difficile* isolate, ribotype 027 (RT027) human isolate 431, and followed the mice for the development of severe outcomes of moribundity and histopathologic cecal tissue damage. Since the murine host and *C. difficile* isolate were the same and only the gut community varied, the variation in disease severity we observed was attributable to the microbiome.

@@

***C. difficile* is able to infect germ-free mice colonized with human fecal microbial communities without perturbation.**  To produce gut microbiomes with greater variation than those found in conventional mouse colonies, we colonized germ-free mice with bacteria from human feces [@Nagpal2018]. We inoculated germ-free C57BL/6 mice with homogenized feces from one of 15 human fecal samples via oral gavage. The human fecal samples were selected because they represented diverse community structures based on community clustering [@Schubert2014]. The gut communities were allowed to establish themselves for two weeks post inoculation [@Gillilland2012]. We then surveyed the bacterial members of the gut communities by 16S rRNA gene sequencing of fecal pellets (Figure 1A). The bacterial communities from each mouse grouped more closely to those communities from mice that received the same human fecal donor community than to the mice who received a different human fecal donor community (Figure 1B). The communities were primarily composed of populations of *Clostridia*, *Bacteroidia*, *Erysipelotrichia*, *Bacilli*, and *Gammaproteobacteria*. However, the gut bacterial communities of each donor group of mice harbored unique relative abundance distributions of the shared bacterial classes.

Next, we tested this set of mice with their human-derived gut microbial communities for susceptibility to *C. difficile* infection. A typical mouse model of CDI requires pre-treatment with antibiotics, such as clindamycin, to become susceptible to *C. difficile* colonization [@Chen2008; @Schubert2015]. However, we wanted to avoid modifying the communities with an antibiotic to maintain their unique microbial compositions and ecological relationships. Since these communities came from people who had increased risk of CDI, such as hospitalization or recent antibiotic use [@Schubert2014], we decided to test if *C. difficile* was able to infect these mice without any antibiotic perturbation. We hypothesized that *C. difficile* would be able to colonize a subset of mice whose communities were susceptible from perturbations either the donor experienced or from the ecological selection when colonizing the murine gut. Mice were challenged with 10$^{3}$ *C. difficile* RT027 human isolate 431 spores. The mice were followed for 10 days post-challenge and their stool was collected and plated for *C. difficile* colony forming units (CFU) to determine the extent of the infection. Surprisingly, mice from all donors were colonized (Figure 2). Only two mice were able to resist *C. difficile* infection and both received their community from the same donor, N3. By colonizing germ-free mice with different human fecal communities, we were able to generate diverse gut communities in mice, which were susceptible to *C. difficile* infection without further modification of the gut community.

**Infection severity varies by initial community.** After we challenged the mice with *C. difficile*, we investigated the outcome from the infection and its relationship to the initial community. We followed the mice for 10 days post-challenge for colonization density, toxin production, and mortality. Seven mice, from Donors N1, N2 and N5, were not colonized on the day after *C. difficile* challenge but were infected (>10$^{6}$) by the end of the experiment. All mice that received their community from Donor M1 through M6 succumbed to the infection and became moribund within 3 days post-challenge. The remaining mice, except the uninfected Donor N3 mice, maintained *C. difficile* infection through the end of the experiment (Figure 2). At 10 days post-challenge, or earlier for the moribund mice, mice were euthanised and fecal pellets were assayed for toxin activity and cecal tissue was collected and scored for histopathologic signs of disease (Figure 3). Overall, there was greater toxin activity detected in the stool of the moribund mice (*P* = 0.003). However, when looking at each group of mice, we observed a range in toxin activity for both the moribund and non-moribund mice (Figure 3A). Non-moribund mice from Donors N4 through N9 had comparable toxin activity as the moribund mice. Additionally, not all moribund mice had toxin activity detected in their stool. This discordance between toxin activity and disease severity has been previously described [@Akerlund2006; @Vitucci2020]. Next we examined the cecal tissue for histopathologic damage. Moribund mice had high levels of epithelial damage, tissue edema, and inflammation (Figure S1) similar to previously reported histopathologic findings for *C. difficile* RT027 [@Cowardin2016]. As observed with toxin activity, the moribund mice had higher histopathologic scores than the non-moribund mice (*P* < 0.001). However, unlike the toxin activity, all moribund mice had consistently high histopathologic summary scores (Figure 3B). The non-moribund mice, Donor groups N1 through N9, had a range in tissue damage from none detected to similar levels as the moribund mice, which grouped by community donor. Together, the toxin activity, histopathologic score, and moribundity showed variation across the donor groups but were largely consistent within each donor group.

**Microbial community members explain variation in CDI severity.** We next interrogated the bacterial communities at the time of *C. difficile* challenge (day 0) for their relationship to infection outcomes using linear discriminant analysis (LDA) effect size (LEfSe) analysis to identify individual bacterial populations that could explain the variation in disease severity. We split the mice into groups by severity level based on their moribundity and histopathologic score. We dichotomized the histopathologic scores into high and low groups by splitting on the median score. This analysis revealed 20 genera that were significantly different by the disease severity (Figure 4A). Bacterial genera *Turicibacter*, *Streptococcus*, *Staphylococcus*, *Pseudomonas*, *Phocaeicola*, *Parabacteroides*, *Bacteroides*, and *Escherichia/Shigella* were detected at higher relative abundances in the mice that became moribund. Populations of *Anaerotignum*, *Coprobacillus*, *Enterocloster*, and *Murimonas* were more abundant in the non-moribund mice that would develop only low intestinal injury. To understand the role of toxin activity in disease severity, we applied LEfSe to identify the genera most likely to explain the observed differences in toxin activity (Figure 4B). Many genera that associated with toxin production were also associated with moribundity, such as populations of *Escherichia/Shigella* and *Bacteroides*. Likewise, there were genera such as *Anaerotignum*, *Enterocloster*, and *Murimonas* that were associated with no toxin production that also exhibited greater relative abundance in communities from non-moribund mice with a low histopathologic score. Lastly, we tested for correlations between the endpoint relative abundances of bacterial operational taxonomic units (OTUs) and the histopathologic summary score (Figure 4C). The endpoint relative abundance of *Bacteroides* was positively correlated with histopathologic score, as its day 0 relative abundance did with disease severity (Figure 4A). Populations of *Klebsiella* and *Prevotellaceae* were positively correlated with the histopathologic score and were increased in the group of mice with detectable toxin. This analysis identified bacterial genera that were associated with the variation in moribundity, histopathologic score, and toxin activity of CDI.

We next determined whether collectively, bacterial community membership and relative abundance could be predictive of the CDI disease outcome. We trained random forest models with bacterial community relative abundance data from the day of colonization at each taxonomic rank to predict toxin production, moribundity, and day 10 post-challenge histopathologic summary score. For predicting if toxin would be produced, microbial populations aggregated by phylum rank classification performed similarly as models using lower taxonomic ranks (AUROC = 0.83, Figure S2). *C. difficile* was more likely to produce toxin when the community it infected had less abundant populations of *Verrucomicrobia* and *Campilobacterota* and had more abundant populations of *Proteobacteria* (Figure 5A). Next, we assessed the ability of the community to predict moribundity. Bacteria grouped by class rank classification was sufficient to predict which mice would succumb to the infection before the end of the experiment (AUROC = 0.91, Figure S2). The features with the greatest effect showed that communities with greater populations of bacteria belonging to *Bacilli* and *Firmicutes* and reduced populations of *Erysipelotrichia* were more likely to result in moribundity (Figure 5B). Only one other class of bacteria was decreased in moribund mice, a group of unclassified *Clostridia*. Lastly, the relative abundances of genera were able to predict a high or low histopathologic score (histopathologic scores were dichotomized as in previous analysis, AUROC = 0.99, Figure S2). No genera had a significantly greater effect on the model performance than any others, indicating the model was reliant on many genera for the correct prediction. The model used some of the genera identified in the LEfSe analysis, such as *Coprobacillus*, *Anaerostipes*, and *Hungatella*. Communities with greater abundances of *Hungatella*, *Eggerthella*, *Bifidobacterium*, *Duncaniella* and *Neisseria* were more likely to have high histopathologic scores. These models have shown that the relative abundance of bacterial populations and their relationship to each other could be used to predict the variation in moribundity, histopathologic score, and toxin activity of CDI.

**Disease severity consistent by donor community across isolates.** Since Carlson et al. showed *C. difficile* isolates with lower germination rates were more likely to result in severe disease [@Carlson2015], we next tested whether the effect of the community on CDI outcomes was conserved across *C. difficile* isolates with higher rates of germination. We colonized mice with human fecal slurry of either Donor N7, N6, or M4 and allowed two weeks for the community to establish. We then challenged the mice with a different *C. difficile* isolate and all mice were infected. The moribundity and histopathologic score outcomes when challenged with a different *C. difficile* human isolate were comparable to the outcomes of challenges with *C. difficile* RT027 human isolate 431 by their donor group (Figure 6). For the isolates and communities we tested, the effect of the gut bacterial community on disease severity was greater than the effect of the variation in these isolates.

@@

Challenging mice colonized with different human fecal communities with a single *C. difficile* isolate, RT027 isolate 431, demonstrated that variation in the gut bacteria affects the *C. difficile* infection disease severity. Our analysis revealed an association between the relative abundance of bacterial community members and disease severity. We used the relative abundance patterns of community members associated with disease outcomes to select donor communities to test whether the disease severity was conserved when challenged with a different *C. difficile* isolate. The disease outcomes from the challenge with the second *C. difficile* isolate matched the outcomes of the challenge with *C. difficile* RT027 isolate 431. Therefore, the effect of the community variation on disease severity *in vivo* was greater than the effect of *C. difficile* isolate variation [@Carlson2013; @Carlson2015]. Overall, these results have demonstrated that the bacterial community contributed to the disease severity of *C. difficile* infection.

CDI severity can range from no symptoms to mild symptoms, such as diarrhea, to severe symptoms, such as toxic megacolon or death. Physicians use prediction tools to identify patients most at risk of developing a severe infection using white blood cell counts, serum albumin level, or serum creatinine level [@Lungulescu2011; @McDonald2018; @Zar2007]. Those levels are driven by the activities in the intestine [@diMasi2018]. Research into the drivers of this variation have revealed factors that make *C. difficile* more virulent. Strains are categorized for their virulence by the presence and production toxin TcdA, toxin TcdB, and binary toxin and the prevalence in outbreaks, such as ribotypes 027 and 078 [@AbernathyClose2020; @Theriot2011; @Goorhuis2008; @OConnor2009; @Chen2008]. However, other studies have shown that virulence is not necessarily linked with toxin production [@Rao2015] or the strain [@Walk2012]. Furthermore, there is variation in the genome, growth rate, sporulation, germination, and toxin production in different isolates of a strain [@Carlson2013; @Carlson2015; @Saund2021; @He2010]. This variation may help explain why severe CDI prediction tools often miss identifying many patients with CDI that will develop severe disease [@Dieterle2020; @Butt2013; @Perry2021; @vanBeurden2017]. Therefore, it is necessary to gain a full understanding of all factors contributing to disease variation to improve our ability to predict severity.

The state of the gut bacterial community determines the ability of *C. difficile* to colonize and persist in the intestine. *C. difficile* is unable to colonize an unperturbed healthy gut community and is only able to become established after a perturbation [@Schubert2015]. Once colonized, the different communities lead to different metabolic responses and dynamics of the *C. difficile* population [@Jenior2017; @Jenior2018; @Lesniak2021]. The gut bacteria metabolize primary bile acids into secondary bile acids [@Staley2016; @Long2017]. The concentration of these bile acids affect the germination, growth, toxin production and biofilm formation [@Sorg2008; @Sorg2010; @Thanissery2017; @Dubois2019]. Members of the bacterial community also affect other metabolites *C. difficile* utilizes. *Bacteroides thetaiotaomicron* produce sialidases which release sialic acid from the mucosa for *C. difficile* to utilize [@Ng2013; @Ferreyra2014]. The nutrient environment affects toxin production [@MartinVerstraete2016]. Thus, many of the actions of the gut bacteria modulate *C. difficile* in ways that could affect the infection and resultant disease.
